Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-08-15 Manuscript/manuscriptPS.tex
Update text

@@

\todo{New figures are delivered by Pavel Buslaev. This should be rewritten.}
None of the tested models reproduce the experimental headgroup and 
glycerol backbone order parameters of DOPS and POPS within the
experimental error bars (Fig. \ref{HGorderParametersPS}). 
The tested models perform generally less well than the models tested
for PC headgroup in the previous study (Fig. 2 in Ref.~\cite{botan15}),
which is also evindent from the comparison between subjective rankings
of the model quality for the headgroup and glycerol backbone 
(Fig. \ref{comparisonTablePS} and Fig. 4 in Ref.~\cite{botan15}).
Therefore, the models cannot be straightforwardly used to interpret the
structural differences between PC and PS headgroups. 
However, the differences are partially reproduced by the two best performing 
models for the $\alpha$ and $\beta$-carbons of PS headgroups, Slipids and CHARMM36.
Both reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order 
in parameter in the PS headgroups (Fig. \ref{HGorderParametersPS} and Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the dihedral angle distributions 
%of CHARMM36 and Slipids in 
in these two models share significant similarities in the headgroup region (Fig.~\ref{dihedralsHG}). 
However, the glycerol backbone order parameters in Slipids model
significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
which can be related to the differences in the dihedral angle distributions 
of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}). 
Similar difference was previously observed for PC lipids \cite{botan15}
and the conformational differences are illustrated in Figure \ref{glycerol_buslaev}. 
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/figS8.png}
  \caption{\label{HGstructuresPS}
    Overlayed snapshots from glycerol backbone and headgroup region from 
    different simulations of PS lipids.
  }
\end{figure*}

@@

Ion binding affinity to PS containing membranes can be most conveniently measured 
from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
not observed \cite{feigenson86,mattai89,roux90,roux91}. Also, the headgroup order 
parameters of PC can be used to detect ion binding affinity to the mixed lipid 
bilayers, see section~\ref{electrometerFORmixtures}. As expected from the previous study of PC lipid bilayers \cite{catte16}, 
the decrease of POPC headgroup order parameters in POPC:POPS mixtures as a function of Ca$^{2+}$ concentration 
is overestimated with respect to experiments \cite{roux90} in almost all the 
tested simulation models (Fig. \ref{changesWITHCaClPS}), indicating overestimated calcium binding
binding affinity. Only exception is the CHARMM36 model with the special NBfix~\cite{kim16} interaction, 
incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
which underestimates the calcium binding affinity. With this interaction, the 
calcium and sodium binding affinities to POPC bilayer are equally weak (see section S7), in contrast to experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the ion binding peaks in
density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}) are underestimated 
in the CHARMM36/NBfix model, but overestimated in the other tested models.

>>>>

\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/figS8.png}
  \caption{\label{HGstructuresPS}
    Overlayed snapshots from glycerol backbone and headgroup region from 
    different simulations of PS lipids.
  }
\end{figure*}

@@

As previously observed for PC lipids, the order parameters and structures
of headgroup and glycerol backbone show wide variety between different simulation
models for PS lipids (Figs. \ref{HGorderParametersPS} and \ref{HGstructuresPS}).
The models for PS lipids perform generally less well than for
PC lipids in the previous study
(Figs. 2 and 4 in Ref.~\cite{botan15} vs. Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS}).
Therefore, interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward. The two best performing models,
Slipids and CHARMM36, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. 2 in Ref.~\citenum{botan15} vs. Fig. \ref{HGorderParametersPS}).
Interestingly, the dihedral angle distributions 
%of CHARMM36 and Slipids in 
in these two models share significant similarities in the headgroup region (Fig.~\ref{dihedralsHG}). 
However, the glycerol backbone order parameters in Slipids model
significantly differ from CHARMM36 results and experiments (Fig. \ref{HGorderParametersPS}),
which can be related to the differences in the dihedral angle distributions 
of C1-C2-C3-O31 and C2-C3-O31-C31 (Fig.~\ref{dihedralsGLY}). 
Similar difference was previously observed for PC lipids \cite{botan15}
and the conformational differences are illustrated in Figure \ref{glycerol_buslaev}.

@@

Ion binding affinity to PS containing membranes can be most conveniently measured 
from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}. In addition, the ion
binding affinity to such mixtures can be detected using the PC lipid headgroup order parameters,
see section~\ref{electrometerFORmixtures}. As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in POPC:POPS (5:1) mixtures 
with respect to the experiments \cite{roux90} (Fig. \ref{changesWITHCaClPS}), indicating overestimated
calcium binding binding affinity. Only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the order parameter changes
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity,
manifested by the peaks in the density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}),
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
in all the other tested models.

### 2018-08-15 Manuscript/manuscriptPSsuppl.tex
Update text

\section{Calcium bindinf to POPC in CHARMM36 simulation with NBfix}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
  The response of headgroup order parameters to the fixed amount of cationic surfactants in
  POPC bilayer is compared between simulations and experiments \cite{scherer89}.}
\end{figure}

>>>>

\section{Calcium binding to POPC in CHARMM36 simulation with NBfix}\label{CHARMMcalciumNBfix}

The response of POPC headgroup order parameters to the CaCl$_2$ concentration
are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 containing
the NBfix for interactions between calicum and lipid oxygens \cite{kim16}
(Fig. \ref{changesWITHCaClPS} in the main text),
indicating that the calcium binding to the bilayer is too weak with these
parameters. Without the NBfix term, the calcium binding affinity to pure POPC
lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.
However, after employing the NBfix term, the response of headgroup order
parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
POPC bilayer are underestimated. Notably, CHARMM36 simulations with the NBfix
terms \cite{venable13,kim16} employed predict similar binding affinity for sodium and calcium.
In conclusion, the special NBfix~\cite{kim16} for calcium, 
incorporated in the parameters give by the CHARMM-GUI at the time
of running the simulations (January 2018), leads to the underestimation of
calcium binding affinity to pure PC and mixed PC:PS bilayers.

\begin{figure}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/OP_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{OP_CHARMM_CaCl_POPC_NBFix}
    Headgroup order parameters from CHARMM36 simulations of POPC mixed with ions having the NBfix term employed for sodium \cite{venable13} ({\it left})
    and calcium \cite{kim16} ({\it right}) compared with the experimental data \cite{akutsu81,altenbach84}.
    Simulation files for CHARMM36 with the NBfix term in calcium are available from Ref. \citenum{CHARMM36withCaClNBfix}.
  }
  \todo{Sources for other simulations by Ricky Nencini.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/density_profile_CHARMM_CaCl_POPC_NBFix.pdf}
  \caption{\label{density_profile_CHARMM_CaCl_POPC_NBFix}
    Density profiles along membrane normal from CHARMM36 simulations with (middle)
    and without (top) the NBfix term for calcium \cite{kim16} compared to the simulation
    with the NBfix term for sodium \cite{venable13} (bottom). The simulation data are the same as in figure \ref{OP_CHARMM_CaCl_POPC_NBFix}.
  }
\end{figure}
